---
title: "Exp35 2015 data analysis"
author: "JBH"
date: "May 23, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(asreml)
library(dplyr)
setwd("M:/File_Server/FieldBooks/Data15/Exp35")
```

## Get ear rot, fungal density, and fumonisin data

I noticed a couple of plots with no ear counts and no ear rot values, but Laura had fumonisin data, I am knocking those out.
```{r getData}
dat35 = read.csv("Exp35_2015_rot_density_fum.csv")
dat35 = mutate(dat35, Fum_LM = FUMppb/1000, 
               Rep = as.factor(Rep), Block = as.factor(Block), 
               Fum_LM = ifelse(is.na(NoEars), NA, Fum_LM)) %>% rename(Fum_TM = fum_ppm_TPM) %>% select(Rep,Block,Plot,Entry,InoculationMethod, NoEars, AvgRot, Fum_LM, Fum_TM)
tbl_df(dat35)
```

Check histograms of raw data
```{r}
for (trait in c("AvgRot", "Fum_LM", "Fum_TM")){
  hist(dat35[,trait], main = trait)
}
```


Compare Laura vs. Thiago ppm overall and each against mean ear rot
```{r cor}
cor(dat35[,c("AvgRot", "Fum_LM", "Fum_TM")], use = "pairwise.complete.obs")
```
Neither fumonisin value is very highly correlated with ear rot, but fum_ppm_TPM is a little better.

```{r pairs}
pairs(dat35[,c("AvgRot", "Fum_LM", "Fum_TM")], main = "Raw data relationships\nincluding both inoculation methods")
```

## Compare toothpick inoculations to backpack inoculations

```{r long_wide}
dat35.syr = filter(dat35, InoculationMethod == "Syringe") %>%
  rename(No_Ears_syr = NoEars, Rot_syr = AvgRot, Fum_LM_syr = Fum_LM, Fum_TM_syr = Fum_TM) %>% select(-InoculationMethod)

dat35.tooth = filter(dat35, InoculationMethod == "Toothpick") %>% 
  rename(No_Ears_tooth = NoEars, Rot_tooth = AvgRot, Fum_LM_tooth = Fum_LM, Fum_TM_tooth = Fum_TM) %>% select(-InoculationMethod)

dat35.wide = merge(dat35.syr, dat35.tooth, all = T)
```

Check the correlations and plot them
```{r}
cor(select(dat35.wide, -c(Rep, Block, Plot, Entry, No_Ears_syr, No_Ears_tooth)), use = "pairwise.complete.obs")
```
The best correlations are between the two fumonisin measures with toothpick inoculation or with syringe, but not across methods!
```{r, fig.width=8, fig.height=9}
pairs(select(dat35.wide, -c(Rep, Block, Entry, No_Ears_syr, No_Ears_tooth)), main = "Raw data relationships, \nseparating inoculation methods")
```
It really looks like outliers are driving these relationships.

##Ear rot raw data analysis

Make a function to generate initial analysis results on raw data.
Note the use of assign() here, this is to work around a problem referencing form object later in the function call. 

see this:
https://www.vsni.co.uk/forum/viewtopic.php?t=1081


```{r}
exp35_lmm = function(data, trait){
  form = as.formula(paste(trait, "~ Entry*InoculationMethod"))
  assign("form", form, envir = globalenv() )
  mod = asreml(fixed = form, random = ~ Rep + Block:Rep, weights = NoEars, data = data, na.method.X='omit')

  print(summary(mod))
  print(wald.asreml(mod, denDF = "default", ssType = "conditional", data = data))
  plot(mod)
  
  return(mod)
}

rot.raw = exp35_lmm(dat35, 'AvgRot')
```
## Analysis of raw fumonisin Laura data

```{r}
Fum_LM.raw = exp35_lmm(dat35, 'Fum_LM')
```

## Analysis of raw fumonisin Thiago data
```{r}
Fum_TM.raw = exp35_lmm(dat35, 'Fum_TM')
```


Looks like we need a transformation of the data. Maybe a log transformation will help all of the traits
```{r}
dat35 = mutate(dat35, 
               ln_Rot = log(1 + AvgRot),
               ln_Fum_LM = log(1 + Fum_LM),
               ln_Fum_TM = log(1 + Fum_TM))
```

Check the correlations among transformed values, across inoculation methods.
```{r}
cor(dat35[,c("ln_Rot", "ln_Fum_LM", "ln_Fum_TM")], use = "pairwise.complete.obs")
```
```{r}
pairs(dat35[,c("ln_Rot", "ln_Fum_LM", "ln_Fum_TM")])
```
OK! this looks much nicer!

Check the correlations among transformed values separately for each inoculation method.

```{r}
dat35.wide = mutate(dat35.wide, 
                    ln_Rot_syr = log(1 + Rot_syr),
                    ln_Fum_LM_syr = log(1 + Fum_LM_syr),
                    ln_Fum_TM_syr = log(1 + Fum_TM_syr),
                    ln_Rot_tooth = log(1 + Rot_tooth),
                    ln_Fum_LM_tooth = log(1 + Fum_LM_tooth),
                    ln_Fum_TM_tooth = log(1 + Fum_TM_tooth)
                    )

cor(dat35.wide[,c("ln_Rot_syr", "ln_Fum_LM_syr", "ln_Fum_TM_syr", "ln_Rot_tooth","ln_Fum_LM_tooth", "ln_Fum_TM_tooth")], use = "pairwise.complete.obs")
```


## Analysis of log transformed Ear Rot

```{r}
Rot.ln = exp35_lmm(dat35, 'ln_Rot')
```
## Analysis of log transformed fumonisin (Laura)

```{r}
Fum_LM.ln = exp35_lmm(dat35, 'ln_Fum_LM')
```

## Analysis of log transformed fumonisin (Thiago)
```{r}
Fum_TM.ln = exp35_lmm(dat35, 'ln_Fum_TM')
```

Log transformed variables have better residuals distributions, so we will use them. 

#Now compute heritabilities for each combination of inoculation method and trait

Cullis estimator is :
$$
h^2 = 1 - \dfrac{(Avg.\ variance\ of\ line\ comparisons)}{2Vg}
$$
```{r}
exp35_h2 = function(data, inoc, trait){
  data.sub = data[data$InoculationMethod == inoc,]
  form = as.formula(paste(trait, "~ 1"))
  assign("form", form, envir = globalenv() )
  mod = asreml(fixed = form, random = ~ Rep + Block:Rep + Entry, weights = NoEars, data = data.sub, na.method.X='omit', maxiter = 25, trace = F)

  print(summary(mod))

  #estimate heritability with Cullis estimator
  preds = predict(mod, classify = "Entry", data = data)
  avsed = preds$predictions$avsed
  
  Vcomps = summary(mod)$varcomp
  Vg = Vcomps["Entry!Entry.var", "component"]
  h2 =1 -((avsed**2)/(2*Vg))
  print(paste("Heritability of line means for", trait, "and", inoc, ":", round(h2, 2)))
    
  return(h2)
}

```
Apply the heritability estimator over each combination of trait and inoculation method
```{r}
h2.results = list()
for (inoc in c("Syringe", "Toothpick")){
  for (trait in c("ln_Rot", "ln_Fum_LM", "ln_Fum_TM")){
    h2.results[[paste(inoc, trait)]] = exp35_h2(dat35, inoc, trait)
  }
}
```

```{r}
for (i in names(h2.results)) {
  print(i)
  print(paste("heritability of line means =", round(h2.results[[i]], 2)))
}
```
We will also check results for the average values across inoculation methods and assay methods. ln_FUM_avg is the average across inoculation methods and assay methods. NoEars is used to weight the means for the two different inoculation methods and a new NoEars value is computed as the sum of number of ears for the two methods, used for weighting the mixed model analysis below.

```{r}
dat35b = group_by(dat35, Rep, Plot, Entry) %>% 
  summarise(ln_Rot_mn = weighted.mean(ln_Rot, NoEars, na.rm = T),
            ln_Fum_LM_mn = weighted.mean(ln_Fum_LM, NoEars, na.rm = T), 
            ln_Fum_TM_mn = weighted.mean(ln_Fum_TM, NoEars, na.rm = T),
            NoEars = sum(NoEars)) %>%
  mutate(ln_FUM_avg = rowMeans(cbind(ln_Fum_LM_mn, ln_Fum_TM_mn), na.rm = T))
```
This works nicely to get the means:
```{r}
tbl_df(dat35b)
```
But, we lost the incomplete block info, so need to grab that and merge it back in.
```{r}
dat35.blocks = group_by(dat35, Rep, Plot) %>% summarize(Block = unique(Block))
tbl_df(dat35.blocks)
```
```{r}
dat35b = merge(dat35.blocks, dat35b)
```

One more heritability function, this time to work on the average trait values in dat35b data frame
```{r}
exp35b_h2 = function(data, trait){
  form = as.formula(paste(trait, "~ 1"))
  assign("form", form, envir = globalenv() )
  mod = asreml(fixed = form, random = ~ Rep + Block:Rep + Entry, weights = NoEars, data = data, na.method.X='omit', maxiter = 25, trace = F)

  print(summary(mod))
  plot(mod)

  #estimate heritability as the average reliability of line predictions
  preds = predict(mod, classify = "Entry", data = data)
  avsed = preds$predictions$avsed
  
  Vcomps = summary(mod)$varcomp
  Vg = Vcomps["Entry!Entry.var", "component"]
  h2 =1 -((avsed**2)/(2*Vg))
  print(paste("Heritability of line means for", trait, ":", round(h2, 2)))
    
  return(h2)
}
```

```{r}
h2.results.b = list()
for (trait in c("ln_Rot_mn", "ln_Fum_LM_mn", "ln_Fum_TM_mn", "ln_FUM_avg")){
  h2.results.b[[trait]] = exp35b_h2(dat35b, trait)
}
```
Get the summary of heritabilities

```{r}
for (i in names(h2.results.b)) {
  print(i)
  print(paste("heritability of line means =", round(h2.results.b[[i]], 2)))
}
```
Now get the line mean values, output a file of means. 
```{r}
exp35_means = function(data, trait){
  form = as.formula(paste(trait, "~ 1 + Entry"))
  assign("form", form, envir = globalenv() )
  mod = asreml(fixed = form, random = ~ Rep + Block:Rep, weights = NoEars, data = data, na.method.X='omit', maxiter = 25)

  preds = predict(mod, classify = "Entry", data = data)
  BLUEs = preds$predictions$pvals[c("Entry", "predicted.value")]
  
  #get the avg. standard error of line mean comparisons
  avgLSD = data.frame(Entry = "Avg LSD", predicted.value = 2*preds$predictions$avsed)
  
  BLUEs = rbind(BLUEs, avgLSD)
  return(BLUEs)
}
```


```{r}
exp35.means = list()
for (trait in c("ln_Rot_mn", "ln_Fum_LM_mn", "ln_Fum_TM_mn", "ln_FUM_avg")){
  meanies = exp35_means(dat35b, trait)
  names(meanies)[2] = trait
  exp35.means[[trait]] = meanies
}

exp35.means.df = exp35.means[[1]]
for (i in 2:4) {
  exp35.means.df = merge(exp35.means.df, exp35.means[[i]])
}
tbl_df(exp35.means.df)
```
Back-transform the data
```{r}
exp35.means.df = mutate(exp35.means.df,
                        Rot_mn = exp(ln_Rot_mn) - 1,
                        Fum_LM_mn = exp(ln_Fum_LM_mn) - 1,
                        Fum_TM_mn = exp(ln_Fum_TM_mn) - 1,
                        FUM_avg = exp(ln_FUM_avg) - 1)

#force Avg LSD to bottom of data frame
exp35.means.out = exp35.means.df
exp35.means.out$pos = ifelse(exp35.means.out$Entry == "Avg LSD", 2, 1)
exp35.means.out = exp35.means.out[order(exp35.means.out$pos, exp35.means.out$Entry),]
exp35.means.out = exp35.means.out[,c("Entry", "ln_Rot_mn", "Rot_mn", "ln_FUM_avg", "FUM_avg")]
names(exp35.means.out) = c("Entry", "ln_EarRot", "EarRot", "ln_FUM", "FUM")
write.csv(exp35.means.out, file = "Exp35_2015_means.csv", row.names = F)
```

```{r}
cor(filter(exp35.means.df, Entry != "Avg LSD") %>% select(-Entry) )
```

```{r}
pairs(filter(exp35.means.df, Entry != "Avg LSD") %>% select(-Entry), main = "Line mean averages\nover inoculation methods and assays")
```
